---
title: "rfsa: A package for accessing and USDA Farm Service Agency data"
output:
  github_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set( 
  collapse = TRUE, 
  comment = "#>",  
  fig.path = "man/figures/README-", 
  out.width = "100%",
  cache = FALSE  
) 

devtools::load_all()
library(dplyr)
library(kableExtra)
library(knitr)

```

# Introduction
The `rfsa` package provides cleaned and aggregated version of publicly available data sets related to programs administered by the USDA, Farm Service Agency. 

Disclaimer: This product uses data provided by the USDA, but is not endorsed by or affiliated with USDA or the Federal Government.

```{r data_set_table, include=FALSE, cache=FALSE}
  # initialize a dataframe with the columns "dataset","rows","columns","years","description"
  df <- data.frame(dataset = character(),
                   rows = integer(),
                   columns = integer(),
                   years = character(),
                   description = character(),
                   last_updated = character(),
                   data_downlod_date = character(),
                   included_columns = character(),
                   stringsAsFactors = FALSE)
  
  # fill in the dataframe with the dataset names, number of rows, number of columns, years, and description using the files in "./data"
  for(f in list.files("./data", pattern = "\\.rda$")){
    # load the dataset
    load(paste0("./data/", f))
    
    # get the name of the dataset
    name <- gsub("\\.rda$", "", f)
    
    # get the number of rows and columns
    rows <- nrow(get(name))
    columns <- ncol(get(name))
    
    if("program_year" %in% colnames(get(name))){
      years = get(name)$program_year
    } else if ("marketing_year" %in% colnames(get(name))){ 
      years = as.numeric(substr(get(name)$marketing_year, 1, 4))
    } else {
      years = get(name)$year
      if(is.character(years)){
        years = as.numeric(substr(years, 1, 4))
      }
    }
    

    
    # get the years covered by the dataset
    min_year = min(years, na.rm = T)
    min_year = ifelse(min_year < 2014, 2014, min_year)
    max_year = max(years, na.rm = T)
    years <- paste0( min_year, "-", max_year)
    
    # get the dates on the file f
    file_info <- file.info(paste0("./data/", f))
    
    # get the last modified date
    last_modified <- file_info$mtime
    
    # keep only the date part in d/m/y format
    last_modified <- format(last_modified, "%m/%d/%Y")
    
    # find the folder in "/data-raw/fsaArcPlc/input_data" that matches the dataset name
    folder <- list.files(path = "./data-raw/fsaArcPlc/input_data", pattern = name, full.names = TRUE)
  
    # if the folder exists, locate the file that contains `current_year` in the name
    if(length(folder) > 0){
      # find the file corresponding to the most recent year
      current_year_file = NULL
      current_year = 2030
      while(length(current_year_file) == 0 && current_year > 2013){
        # get the file that contains `current_year` in the name
        current_year_file <- list.files(path = folder, pattern = paste0(current_year), full.names = TRUE)
        # decrease the year
        current_year <- current_year - 1
      }
      
      if(length(current_year_file) == 0){
        # if the file does not exist, set the date to NA
        data_downlod_date <- NA
      } else {
        # if the file exists, get the last modified date of the file
        file_info <- file.info(current_year_file)
        
        # get the last modified date
        data_downlod_date <- file_info$mtime
        
        # keep only the date part in d/m/y format
        data_downlod_date <- format(data_downlod_date, "%m/%d/%Y")
      }
    }
      
  
    
    # add a row to the dataframe
    df <- rbind(df, data.frame(dataset = name,
                                rows = rows,
                                columns = columns,
                                years = years,
                                last_updated = last_modified,
                                data_downlod_date = data_downlod_date,
                                included_columns = paste(colnames(get(name)), collapse = ", "),
                                stringsAsFactors = FALSE))
    }

  # define descriptions of the data set
  descriptions_list = list()
  
  # list each file in ./man
  man_files <- list.files(path = "./man", pattern = "\\.Rd$", full.names = TRUE)
  
  # loop through each file
  for(m in man_files){
    # get the name of the dataset
    name <- gsub("\\.Rd$", "", basename(m))
    
    # Read the .Rd file
    rd_lines <- readLines(m)
    
    # Extract the line with the \title{} field
    title_line <- grep("^\\\\title\\{", rd_lines, value = TRUE)
    
    # Remove the LaTeX command and extract the content
    title_text <- sub("^\\\\title\\{(.*)\\}$", "\\1", title_line)
  
    # add the description to the list
    descriptions_list[[name]] <- title_text
  }

  # merge the descriptions into the dataframe
  df <- df %>%
    left_join(data.frame(dataset = names(descriptions_list),
                         description = unlist(descriptions_list)),
              by = "dataset") %>%
    mutate(description = ifelse(is.na(description), "No description available", description))
  

# rename df columns
colnames(df) <- c("Dataset", "Rows", "Columns", "Years", "Last Updated", "Data Download Date", "Included Columns" ,"Description")

# reorder the columns
df <- df[,c("Dataset", "Description", "Rows", "Years", "Last Updated", "Data Download Date", "Included Columns")]

```
# Installation
`rfsa` can be installed directly from GitHub using `remotes::install_github("dylan-turner25/rfsa", force = TRUE)`.

# Usage
The`rfsa` package provides several ways to access data depending on the type of data set. The first way is by loading pre-built data sets using the `data()` function.  For example, to load the `fsaMyaPrice` data set, use the following code.
```{r}
# load the rfsa package
library(rfsa)

# load marketing year average prices
data("fsaMyaPrice")
 
head(fsaMyaPrice)

```

The other way to ob


# ARC and PLC Program Data
The following table reports the currently available data sets related to ARC and PLC in the `rfsa` package. These data sets are cleaned and compiled versions of data available on FSA's [ARC and PLC data landing page](https://www.fsa.usda.gov/resources/programs/arc-plc/program-data). Loading a data set is done using the `data` function. For example, to load the `fsaMyaPrice` data set, use the following code.
```{r}
# load the rfsa package
library(rfsa)

# load marketing year average prices
data("fsaMyaPrice")
 
head(fsaMyaPrice)

```

The following table provides some information about the ARC and PLC specific data sets in the `rfsa` package. The `Last Updated` column indicates the last time the data set was updated in the package. The `Data Download Date` indicates the date that the raw input data for the current year was downloaded from the FSA website (this is important as these data sets are sometimes revised or contain projected values for in progress marketing years).

```{r, include = TRUE, echo = FALSE, cache=FALSE}
 knitr::kable(df) 
``` 

# FSA Individual Payment Files

The USDA Farm Service Agency provides access to [individual payment files](https://www.fsa.usda.gov/tools/informational/freedom-information-act-foia/electronic-reading-room/frequently-requested/payment-files) that contain payment information for programs administered by FSA. The data in these files can be accessed using the `get_fsa_payments()` function. This function pulls data from [pre-cleaned files](https://github.com/dylan-turner25/rfsa/releases/tag/v0.1.0) that are stored as GitHub Releases on the `rfsa` GitHub Repository. This approach minimizes memory overhead and compute time when only a small portion of the data is needed. The `get_fsa_payments()` function has several arguments that allow the user to filter the data. The `year` argument specifies the year of interest (can be a vector of multiple year), the `program` argument specifies the program of interest (ex: "ARC-CO","ARC-IC","PLC", "CRP"), and the `year_type` argument specifies whether to pull data that has been aggregated by `program_year` (i.e. the year corresponding to the event that prompted the payment), the `fiscal_year` (i.e. the Government fiscal year corresponding to the payment), or `payment_year` (i.e. the actual calendar year when funds were disbursed) . The `aggregation` argument specifies whether to aggregate the data at the `national`,`state`, `county`, or `individual` level. The first time that a combination of `year`, `program`, and `year_type` are specified will prompt the relevant files to be downloaded which will then be cached on your local machine. This means that depending on the quanity of data needed, the initial query may take some time, however, subsequent function calls for the same data will be much quicker. To remove all cached data use `clear_rfsa_cache()`. 
Below are several examples of how to use the `get_fsa_payments()` function.

```{r, include = TRUE, echo = TRUE, cache=FALSE}

library(rfsa)

# get national level data on payments made through the conservation reserve program in program year 2023. 
data <- get_fsa_payments(year = 2023, 
                         program = c("CRP"), 
                         year_type = "program", 
                         aggregation = "national")


```


The following table provides a list of all the program abbreviations that can be passed to the `program` argument in the `get_fsa_payments()` function. A complete list of every unique accounting description that occurs in all FSA individual payment files and how each description was classified into one of the program listed below, see [this file](https://github.com/dylan-turner25/rfsa/blob/main/data-raw/fsaFarmPayments/supplementary_files/program_details.csv) (pointing out any suspected incorrect classifications is encouraged!). 
```{r, echo = FALSE, cache=FALSE, include = FALSE}

abb <- distinct(readr::read_csv("./data-raw/fsaFarmPayments/supplementary_files/program_details.csv") %>%
  select(prog_abb, prog_full)) %>%
  rename(program_abbreviation = prog_abb,
         program_full_name = prog_full) 


```

```{r, echo = FALSE}
knitr::kable(abb)
```




# Data Validation Checks

The following table contains data validation checks. These are comparisons between values derived from the `rfsa` package functions against the same values obtained from another source. For example, the first row calculates total ARC-CO payments in program year 2023 using the `get_fsa_payments()` function and compares it to the total ARC-CO payments in program year 2023 from an aggregated file on the FSA website. The `check_passed` column indicates whether the difference between the two values is less than 1%. If you are reading this and have a value of interest that you would like to see added to this table, please open an issue on the GitHub repository with the relevant information including code to generate the value using the `rfsa` package as well as an external source to validate the value against.

```{r, echo = F}

# create a data frame with columns for dataset name, code, package generated value, external value, external source
df <- data.frame(dataset = character(),
                 code = character(),
                 package_value = numeric(),
                 external_value = numeric(),
                 external_source = character(),
                 stringsAsFactors = FALSE)

# add a row for ARC-CO payments in program year 2023
metric = as.numeric(get_fsa_payments(year = 2023,program = c("ARC-CO"),year_type = "program",aggregation = "national")[,"payment_amount"])
external_value = 461724994.11
percentage_difference = (metric - external_value)/(external_value) 
df <- rbind(df, data.frame(value = "National ARC-CO payments in program year 2023",
                             code = 'get_fsa_payments(year = 2023,program = c("ARC-CO"),year_type = "program",aggregation = "national")[,"payment_amount"]',
                             package_value = metric,
                             external_value = external_value,
                             external_source = "https://www.fsa.usda.gov/sites/default/files/2025-01/ARCCO%20Non-ProgYr%20Specific%20Payment%20Data%20%282025-01-06%29.xlsx",
                             percentage_difference = paste0("%",round((percentage_difference * 100),4)),
                             check_passed = ifelse(abs(percentage_difference) < .01,
                                                    cell_spec("✓",   
                                                    format    = "html",
                                                    color     = "white",
                                                    background= "forestgreen",
                                                    bold      = TRUE),
                                                    # red box, white cross
                                                    cell_spec("✗",
                                                    format    = "html",
                                                    color     = "white",
                                                    background= "firebrick",
                                                    bold      = TRUE)),
                             stringsAsFactors = FALSE))


# add a row for ARC-CO payments in program year 2023 using the ARC/PLC program data
data("fsaArcPlcPayments") 
metric = fsaArcPlcPayments %>%
filter(program == "ARC-CO", program_year == 2023) %>%
group_by(program_year) %>%
summarize(payments = sum(payments))
metric = metric$payments
external_value = 461724994.11
percentage_difference = (metric - external_value)/(external_value) 
df <- rbind(df, data.frame(value = "National ARC-CO payments in program year 2023",
                             code =  'data(fsaArcPlcPayments); fsaArcPlcPayments %>% filter(program == "ARC-CO", program_year == 2023) %>% group_by(program_year) %>% summarize(payments = sum(payments))',
                             package_value = metric,
                             external_value = external_value,
                             external_source = "https://www.fsa.usda.gov/sites/default/files/2025-01/ARCCO%20Non-ProgYr%20Specific%20Payment%20Data%20%282025-01-06%29.xlsx",
                             percentage_difference = paste0("%",round((percentage_difference * 100),4)),
                             check_passed = ifelse(abs(percentage_difference) < .01,
                                                    cell_spec("✓",   
                                                    format    = "html",
                                                    color     = "white",
                                                    background= "forestgreen",
                                                    bold      = TRUE),
                                                    # red box, white cross
                                                    cell_spec("✗",
                                                    format    = "html",
                                                    color     = "white",
                                                    background= "firebrick",
                                                    bold      = TRUE)),
                             stringsAsFactors = FALSE))




knitr::kable(df)


```

# Example Usage

## Plot payments made via the Conservation Reserve Program relative to total payments over time 
```{r, cache=TRUE, echo=TRUE, include=TRUE}
library(rfsa)
library(ggplot2)

# note this is aggregating roughly 60 million individual payments behind the scenes
# and takes up roughly 20 GB of memory. A more memory efficient approach would be
# to loop over years and aggregate each year individually to avoid loading all the data into memory at once.
data <- get_fsa_payments(year = 2004:2023,
                         year_type = "program",
                         aggregation = "national")


data %>%
  mutate(program_category = if_else(program_abb %in% c("CRP"),
                                    "Conservation Reserve Program",
                                    "Other")) %>%
  ggplot(aes(x = year, y = payment_amount/1e9, fill = program_category)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Conservation Reserve Program" = "forestgreen",
                               "Other" = "grey")) +
  labs(
    title = "FSA Payments by Program (2004-2023)",
    x = "Program Year",
    y = "Total Payments (Billions USD)",
    fill = "Program"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))


```



## Plot county level payments made through the livestock indemnity program in program year 2023
```{r, cache=TRUE, echo=TRUE, include=TRUE}
library(ggplot2)
library(maps)
library(mapproj)
library(dplyr)

data <- get_fsa_payments(year = 2023, 
                         year_type = "program",
                         program = c("LIP"),
                         aggregation = "county") %>%
  mutate(
    state_fips = substr(fips_fsa, 1, 2),
    county_fips = substr(fips_fsa, 3, 5)
  )

# Get county map data
counties <- map_data("county")

# Get state and county names from FIPS codes
fips_codes <- data %>%
  select(fips_fsa, county_name_fsa) %>%
  distinct() %>%
  mutate(
    state = state.fips$polyname[match(state_cd_fsa, state.fips$fips)],
    county = tolower(county_name_fsa)
  )

# Join payment data with map data
map_data <- counties %>%
  left_join(
    data %>%
      left_join(fips_codes, by = "fips_fsa") %>%
      select(state, county, payment_amount),
    by = c("region" = "state", "subregion" = "county")
  )


# Create the map
ggplot(map_data, aes(x = long, y = lat, group = group, fill = payment_amount)) +
geom_polygon(color = "white", size = 0.1) +
coord_map("albers", lat0 = 30, lat1 = 40) +
scale_fill_viridis_c(
  option = "magma",
  name = "",
  trans = "log10",
  labels = scales::dollar_format(),
  na.value = "grey90",
  direction = -1
) +
labs(
  title = "Total Livestock Indemnity Program Payments in Program Year 2023",
  caption = "Source: FSA Payment Data"
) +
theme_minimal() +
theme(
  axis.text = element_blank(),
  axis.title = element_blank(),
  panel.grid = element_blank()
)


```

