---
title: "Calculating Price Loss Coverage (PLC) Payments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Price Loss Coverage (PLC) Payments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  error = TRUE
)
```

```{r setup}
library(rfsa)
```

## Introduction

The **rfsa** package provides tools for calculating Price Loss Coverage (PLC) payments administered by the USDA Farm Service Agency. PLC is a commodity program that provides payments to farmers when the effective reference price exceeds the higher of the national marketing loan rate or the Marketing Year Average (MYA) price.

This vignette demonstrates how to use the `calc_plc_payment()` function to calculate PLC payments for various crops and scenarios.

## Basic PLC Payment Calculations

### Understanding the Payment Formula

The PLC payment is calculated using this formula:

**Payment = (Reference Price - Comparison Price) × PLC Yield × Base Acres × Coverage Level**

Where:
- **Reference Price** = Effective Reference Price (ERP) for recent years, or Statutory Reference Price (SRP) for 2014-2019
- **Comparison Price** = Higher of MYA Price or National Marketing Loan Rate (NMLR)

### Simple Example with Fixed Values

Let's start with a basic calculation using known values:

```{r basic-example}
# Calculate PLC payment with specified parameters
payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 3.50,
  erp = 4.30,
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

print(paste("PLC payment for 100 acres of corn:", round(payment, 2)))
```

### Breaking Down the Calculation

```{r components-example}
# Example showing the calculation components
crop <- "corn"
erp <- 4.30
mya_price <- 3.50
nmlr <- 2.50
plc_yield <- 180
base_acres <- 100
coverage_level <- 0.85

# Step 1: Determine if payment is due
payment_due <- erp > mya_price
print(paste("Payment due (ERP > MYA)?", payment_due))

# Step 2: Determine comparison price
comparison_price <- max(mya_price, nmlr)
print(paste("Comparison price (max of MYA and NMLR):", comparison_price))

# Step 3: Calculate payment rate per bushel
payment_rate <- erp - comparison_price
print(paste("Payment rate per bushel:", payment_rate))

# Step 4: Calculate total payment
total_payment <- payment_rate * plc_yield * base_acres * coverage_level
print(paste("Total PLC payment:", round(total_payment, 2)))

# Verify with function
function_result <- calc_plc_payment(
  crop = crop,
  program_year = 2024,
  base_acres = base_acres,
  mya_price = mya_price,
  erp = erp,
  nmlr = nmlr,
  plc_yield = plc_yield,
  cov_lvl = coverage_level,
  quiet = TRUE
)

print(paste("Function result matches:", round(function_result, 2) == round(total_payment, 2)))
```

## Working with Different Scenarios

### When No Payment is Due

```{r no-payment-scenario}
# Scenario where ERP <= MYA Price (no payment)
no_payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 4.50,  # High MYA price
  erp = 4.30,        # Lower ERP
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

print(paste("Payment when ERP <= MYA:", no_payment))
```

### When MYA Price is Very Low

```{r low-mya-scenario}
# Scenario where MYA Price <= NMLR (uses NMLR in calculation)
low_mya_payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 2.00,  # Very low MYA price
  erp = 4.30,
  nmlr = 2.50,       # Higher than MYA
  plc_yield = 180,
  quiet = TRUE
)

print(paste("Payment when MYA <= NMLR:", round(low_mya_payment, 2)))
```

## Different Coverage Levels

```{r coverage-levels}
# Compare different coverage levels
payment_85 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 3.50,
  erp = 4.30,
  nmlr = 2.50,
  plc_yield = 180,
  cov_lvl = 0.85,
  quiet = TRUE
)

payment_88 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 3.50,
  erp = 4.30,
  nmlr = 2.50,
  plc_yield = 180,
  cov_lvl = 0.88,
  quiet = TRUE
)

print(paste("85% coverage:", round(payment_85, 2)))
print(paste("88% coverage:", round(payment_88, 2)))
print(paste("Difference:", round(payment_88 - payment_85, 2)))
```

## Using Custom Price Data

You can provide your own MYA price data for ERP calculations:

```{r custom-prices}
# Provide 5 years of MYA price data for custom ERP calculation
custom_mya <- list(
  years = c(2019, 2020, 2021, 2022, 2023),
  price = c(3.60, 3.50, 5.80, 6.80, 4.85)
)

custom_payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = custom_mya,
  srp = 4.30,  # Statutory reference price
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

print(paste("Payment with custom MYA data:", round(custom_payment, 2)))
```

## Working with Real FSA Data

When you don't specify prices and yields, the function will attempt to use FSA datasets:

```{r fsa-data-example}
# Try to use FSA data (may not work if datasets aren't available)
tryCatch({
  # Using FSA data for 2019 (historical year with known data)
  fsa_payment <- calc_plc_payment(
    crop = "corn",
    program_year = 2019,
    base_acres = 100,
    quiet = TRUE
  )
  
  print(paste("Payment using FSA data:", round(fsa_payment, 2)))
}, error = function(e) {
  print("FSA datasets may not be fully available in this environment")
  print("In a full installation, this would use real FSA price and yield data")
})
```

## Crop Types and Location Examples

### Different Crop Types

```{r crop-types}
# Examples with crop types (using fixed values to ensure they work)
rice_long <- calc_plc_payment(
  crop = "rice",
  crop_type = "long grain",
  program_year = 2024,
  base_acres = 150,
  mya_price = 14.50,
  erp = 15.20,
  nmlr = 6.50,
  plc_yield = 7.5,
  quiet = TRUE
)

chickpeas_large <- calc_plc_payment(
  crop = "chickpeas",
  crop_type = "large",
  program_year = 2024,
  base_acres = 75,
  mya_price = 22.00,
  erp = 23.97,
  nmlr = 11.72,
  plc_yield = 2.3,
  quiet = TRUE
)

print(paste("Rice (long grain) payment:", round(rice_long, 2)))
print(paste("Chickpeas (large) payment:", round(chickpeas_large, 2)))
```

### Location-Specific Examples

```{r location-examples}
# Examples using location parameters with FSA data
tryCatch({
  # Try county-specific calculation
  iowa_payment <- calc_plc_payment(
    crop = "corn",
    program_year = 2019,  # Use 2019 as it has more stable data
    base_acres = 100,
    state = "Iowa",
    county = "Story",
    quiet = TRUE
  )
  
  print(paste("Iowa (Story County) payment:", round(iowa_payment, 2)))
}, error = function(e) {
  print("Location-specific data requires full FSA datasets")
  print("In practice, you would specify: state='Iowa', county='Story'")
})
```

## Historical vs. Recent Program Years

```{r historical-comparison}
# Compare 2019 (uses SRP) vs 2024 (uses ERP) calculations
# Using fixed values to demonstrate the difference

# 2019 calculation (would typically use SRP)
payment_2019 <- calc_plc_payment(
  crop = "corn",
  program_year = 2019,
  base_acres = 100,
  mya_price = 3.61,
  srp = 3.70,  # Statutory Reference Price used for 2019
  nmlr = 2.20,
  plc_yield = 180,
  quiet = TRUE
)

# 2024 calculation (uses ERP)
payment_2024 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 3.61,
  erp = 4.30,  # Effective Reference Price used for 2024
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

print(paste("2019 payment (SRP-based):", round(payment_2019, 2)))
print(paste("2024 payment (ERP-based):", round(payment_2024, 2)))
```

## Batch Processing Example

```{r batch-processing}
# Example of calculating payments for multiple scenarios
scenarios <- data.frame(
  scenario = c("High ERP", "Low MYA", "No Payment", "Large Farm"),
  crop = c("corn", "corn", "corn", "soybeans"),
  base_acres = c(100, 100, 100, 500),
  mya_price = c(3.50, 2.00, 4.50, 11.50),
  erp = c(4.30, 4.30, 4.00, 12.95),
  nmlr = c(2.50, 2.50, 2.50, 6.20),
  plc_yield = c(180, 180, 180, 50),
  stringsAsFactors = FALSE
)

# Calculate payments
scenarios$payment <- 0
for (i in 1:nrow(scenarios)) {
  scenarios$payment[i] <- calc_plc_payment(
    crop = scenarios$crop[i],
    program_year = 2024,
    base_acres = scenarios$base_acres[i],
    mya_price = scenarios$mya_price[i],
    erp = scenarios$erp[i],
    nmlr = scenarios$nmlr[i],
    plc_yield = scenarios$plc_yield[i],
    quiet = TRUE
  )
}

print("Batch Processing Results:")
print(scenarios[, c("scenario", "crop", "base_acres", "payment")])

total_payments <- sum(scenarios$payment)
print(paste("Total payments across all scenarios:", round(total_payments, 2)))
```

## Quiet Mode for Production Use

```{r quiet-mode}
# Demonstrate quiet mode for suppressing warnings
# This would normally show warnings about averaging across crop types
verbose_result <- calc_plc_payment(
  crop = "rice",  # Multiple crop types available
  program_year = 2024,
  base_acres = 100,
  mya_price = 14.50,
  erp = 15.20,
  nmlr = 6.50,
  plc_yield = 7.5,
  quiet = FALSE  # Will show warnings
)

quiet_result <- calc_plc_payment(
  crop = "rice",  # Same calculation
  program_year = 2024,
  base_acres = 100,
  mya_price = 14.50,
  erp = 15.20,
  nmlr = 6.50,
  plc_yield = 7.5,
  quiet = TRUE   # Suppresses warnings
)

print(paste("Results are identical:", verbose_result == quiet_result))
```



