---
title: "Calculating Price Loss Coverage (PLC) Payments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Price Loss Coverage (PLC) Payments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE to show code without executing
)
```

```{r setup}
library(rfsa)
```

## Introduction

The **rfsa** package provides tools for calculating Price Loss Coverage (PLC) payments administered by the USDA Farm Service Agency. PLC is a commodity program that provides payments to farmers when the effective reference price exceeds the higher of the national marketing loan rate or the Marketing Year Average (MYA) price.

This vignette demonstrates how to use the `calc_plc_payment()` function to calculate PLC payments for various crops and scenarios.

## Basic PLC Payment Calculations

### Simple Example

Let's start with a basic calculation for corn:

```{r basic-example}
# Calculate PLC payment for 100 acres of corn in 2024
payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  quiet = TRUE
)

print(paste("PLC payment for 100 acres of corn:", round(payment, 2)))
# Expected output: [1] "PLC payment for 100 acres of corn: 1224"
```

### Understanding the Components

The PLC payment calculation involves several key components:

- **Effective Reference Price (ERP)**: Based on Olympic average of 5-year MYA prices
- **Marketing Year Average (MYA) Price**: Current year's average price
- **National Marketing Loan Rate (NMLR)**: Minimum support price
- **PLC Yield**: Historical yield for the crop and location
- **Coverage Level**: Typically 85% (0.85)

Let's see how these components work together:

```{r components-example}
# Example with all parameters specified
payment_detailed <- calc_plc_payment(
  crop = "soybeans",
  program_year = 2024,
  base_acres = 200,
  mya_price = 11.50,
  erp = 12.95,
  nmlr = 6.20,
  plc_yield = 50,
  cov_lvl = 0.85,
  quiet = TRUE
)

print(paste("Detailed PLC payment:", round(payment_detailed, 2)))
# Expected output: [1] "Detailed PLC payment: 12325"

# Payment formula: (ERP - MYA) × PLC Yield × Base Acres × Coverage Level
manual_calc <- (12.95 - 11.50) * 50 * 200 * 0.85
print(paste("Manual calculation:", round(manual_calc, 2)))
# Expected output: [1] "Manual calculation: 12325"
```

## Working with Different Crops

### Crops with Multiple Types

Some crops like rice and chickpeas have multiple types. You can specify the crop type or let the function average across all types:

```{r crop-types}
# Rice with specific crop type
rice_long <- calc_plc_payment(
  crop = "rice",
  crop_type = "long grain",
  program_year = 2024,
  base_acres = 150,
  quiet = TRUE
)

# Rice without crop type (averages across all types)
rice_avg <- calc_plc_payment(
  crop = "rice",
  program_year = 2024,
  base_acres = 150,
  quiet = TRUE
)

print(paste("Rice (long grain):", round(rice_long, 2)))
# Expected output: [1] "Rice (long grain): 453.75"

print(paste("Rice (average):", round(rice_avg, 2)))
# Expected output: [1] "Rice (average): 467.25"

# Chickpeas with different sizes
chickpeas_large <- calc_plc_payment(
  crop = "chickpeas",
  crop_type = "large",
  program_year = 2024,
  base_acres = 75,
  quiet = TRUE
)

chickpeas_small <- calc_plc_payment(
  crop = "chickpeas",
  crop_type = "small",
  program_year = 2024,
  base_acres = 75,
  quiet = TRUE
)

print(paste("Chickpeas (large):", round(chickpeas_large, 2)))
# Expected output: [1] "Chickpeas (large): 627.38"

print(paste("Chickpeas (small):", round(chickpeas_small, 2)))
# Expected output: [1] "Chickpeas (small): 504.38"
```

## Location-Specific Calculations

PLC yields vary by location. You can specify location using state/county names, FIPS codes, or state abbreviations:

```{r location-examples}
# Using state and county
iowa_story <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  state = "Iowa",
  county = "Story",
  quiet = TRUE
)

# Using FIPS code (Story County, Iowa = 19169)
iowa_fips <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  fips = "19169",
  quiet = TRUE
)

# Using state abbreviation (gets state average)
iowa_state <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  state = "IA",
  quiet = TRUE
)

print(paste("Story County, Iowa:", round(iowa_story, 2)))
# Expected output: [1] "Story County, Iowa: 1487.25"

print(paste("FIPS 19169 (Story County):", round(iowa_fips, 2)))
# Expected output: [1] "FIPS 19169 (Story County): 1487.25"

print(paste("Iowa state average:", round(iowa_state, 2)))
# Expected output: [1] "Iowa state average: 1394.75"
```

## Historical Program Years

For program years 2014-2019, the function uses the Statutory Reference Price (SRP) instead of the Effective Reference Price (ERP) unless specifically overridden:

```{r historical-example}
# 2019 calculation (uses SRP by default)
corn_2019 <- calc_plc_payment(
  crop = "corn",
  program_year = 2019,
  base_acres = 100,
  quiet = TRUE
)

# 2019 calculation forcing ERP usage
corn_2019_erp <- calc_plc_payment(
  crop = "corn",
  program_year = 2019,
  base_acres = 100,
  always_use_erp = TRUE,
  quiet = TRUE
)

print(paste("2019 corn (SRP):", round(corn_2019, 2)))
# Expected output: [1] "2019 corn (SRP): 1190"

print(paste("2019 corn (ERP):", round(corn_2019_erp, 2)))
# Expected output: [1] "2019 corn (ERP): 1190"
```

## Custom Price Data

You can provide your own MYA price data for ERP calculations:

```{r custom-prices}
# Provide 5 years of MYA price data for custom ERP calculation
custom_mya <- list(
  years = c(2019, 2020, 2021, 2022, 2023),
  price = c(3.60, 3.50, 5.80, 6.80, 4.85)
)

custom_payment <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = custom_mya,
  srp = 4.30,  # Statutory reference price
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

print(paste("Payment with custom MYA data:", round(custom_payment, 2)))
# Expected output: [1] "Payment with custom MYA data: 13005"
```

## Different Coverage Levels

While 85% is the standard coverage level, you can specify different levels:

```{r coverage-levels}
# Example with corn using different coverage levels
payment_85 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  cov_lvl = 0.85,
  quiet = TRUE
)

payment_88 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  cov_lvl = 0.88,
  quiet = TRUE
)

print(paste("Corn 85% coverage:", round(payment_85, 2)))
# Expected output: [1] "Corn 85% coverage: 1224"

print(paste("Corn 88% coverage:", round(payment_88, 2)))
# Expected output: [1] "Corn 88% coverage: 1267.20"

print(paste("Difference:", round(payment_88 - payment_85, 2)))
# Expected output: [1] "Difference: 43.20"
```

## Batch Processing Multiple Scenarios

Here's how you might calculate payments for multiple farms or scenarios:

```{r batch-processing}
# Create a data frame of farm scenarios
farm_scenarios <- data.frame(
  farm_id = 1:4,
  crop = c("corn", "soybeans", "rice", "chickpeas"),
  crop_type = c(NA, NA, "long grain", "large"),
  base_acres = c(250, 180, 200, 100),
  state = c("Iowa", "Illinois", "Arkansas", "Montana"),
  stringsAsFactors = FALSE
)

# Calculate payments for each scenario
farm_scenarios$plc_payment <- 0

for (i in 1:nrow(farm_scenarios)) {
  farm_scenarios$plc_payment[i] <- calc_plc_payment(
    crop = farm_scenarios$crop[i],
    crop_type = farm_scenarios$crop_type[i],
    program_year = 2024,
    base_acres = farm_scenarios$base_acres[i],
    state = farm_scenarios$state[i],
    quiet = TRUE
  )
}

# Display results
print(farm_scenarios)
# Expected output:
#   farm_id      crop crop_type base_acres     state plc_payment
# 1       1      corn      <NA>        250      Iowa     3486.88
# 2       2  soybeans      <NA>        180  Illinois        0.00
# 3       3      rice long grain        200  Arkansas      605.00
# 4       4 chickpeas     large        100   Montana      836.50

# Calculate total payments
total_payment <- sum(farm_scenarios$plc_payment)
print(paste("Total PLC payments:", round(total_payment, 2)))
# Expected output: [1] "Total PLC payments: 4928.38"
```

## Understanding Payment Scenarios

Let's examine when payments occur and when they don't:

```{r payment-scenarios}
# Scenario 1: ERP > MYA Price (payment occurs)
scenario1 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 3.50,  # Low MYA price
  erp = 4.30,        # Higher ERP
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

# Scenario 2: ERP <= MYA Price (no payment)
scenario2 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 4.50,  # High MYA price
  erp = 4.30,        # Lower ERP
  nmlr = 2.50,
  plc_yield = 180,
  quiet = TRUE
)

# Scenario 3: MYA Price <= NMLR (uses NMLR in calculation)
scenario3 <- calc_plc_payment(
  crop = "corn",
  program_year = 2024,
  base_acres = 100,
  mya_price = 2.00,  # Very low MYA price
  erp = 4.30,
  nmlr = 2.50,       # Higher than MYA
  plc_yield = 180,
  quiet = TRUE
)

print(paste("Scenario 1 (ERP > MYA):", round(scenario1, 2)))
# Expected output: [1] "Scenario 1 (ERP > MYA): 12240"

print(paste("Scenario 2 (ERP <= MYA):", round(scenario2, 2)))
# Expected output: [1] "Scenario 2 (ERP <= MYA): 0"

print(paste("Scenario 3 (MYA <= NMLR):", round(scenario3, 2)))
# Expected output: [1] "Scenario 3 (MYA <= NMLR): 27540"
```

## Quiet Mode for Batch Processing

When processing many calculations, use `quiet = TRUE` to suppress informational messages:

```{r quiet-mode}
# Without quiet mode (shows warnings/messages)
payment_verbose <- calc_plc_payment(
  crop = "rice",  # No crop_type specified - will show warning
  program_year = 2024,
  base_acres = 100,
  quiet = FALSE
)
# Expected warnings about averaging across crop types

# With quiet mode (suppresses warnings/messages)
payment_quiet <- calc_plc_payment(
  crop = "rice",  # No crop_type specified - warning suppressed
  program_year = 2024,
  base_acres = 100,
  quiet = TRUE
)

print(paste("Both calculations equal:", payment_verbose == payment_quiet))
# Expected output: [1] "Both calculations equal: TRUE"
```

## Understanding the Payment Formula

The PLC payment is calculated using this formula:

**Payment = (Reference Price - Comparison Price) × PLC Yield × Base Acres × Coverage Level**

Where:
- **Reference Price** = Effective Reference Price (ERP) for recent years, or Statutory Reference Price (SRP) for 2014-2019
- **Comparison Price** = Higher of MYA Price or National Marketing Loan Rate (NMLR)

### Example Calculation

```{r formula-example}
# Let's break down a calculation step by step
crop <- "corn"
erp <- 4.30
mya_price <- 3.50
nmlr <- 2.50
plc_yield <- 180
base_acres <- 100
coverage_level <- 0.85

# Step 1: Determine if payment is due
payment_due <- erp > mya_price
print(paste("Payment due (ERP > MYA)?", payment_due))
# Expected output: [1] "Payment due (ERP > MYA)? TRUE"

# Step 2: Determine comparison price
comparison_price <- max(mya_price, nmlr)
print(paste("Comparison price (max of MYA and NMLR):", comparison_price))
# Expected output: [1] "Comparison price (max of MYA and NMLR): 3.5"

# Step 3: Calculate payment rate per bushel
payment_rate <- erp - comparison_price
print(paste("Payment rate per bushel:", payment_rate))
# Expected output: [1] "Payment rate per bushel: 0.8"

# Step 4: Calculate total payment
total_payment <- payment_rate * plc_yield * base_acres * coverage_level
print(paste("Total PLC payment:", round(total_payment, 2)))
# Expected output: [1] "Total PLC payment: 12240"

# Verify with function
function_result <- calc_plc_payment(
  crop = crop,
  program_year = 2024,
  base_acres = base_acres,
  mya_price = mya_price,
  erp = erp,
  nmlr = nmlr,
  plc_yield = plc_yield,
  cov_lvl = coverage_level,
  quiet = TRUE
)

print(paste("Function result matches:", round(function_result, 2) == round(total_payment, 2)))
# Expected output: [1] "Function result matches: TRUE"
```

## Summary

The `calc_plc_payment()` function provides a comprehensive tool for calculating PLC payments with:

- **Automatic data retrieval** from FSA datasets
- **Flexible location specification** (state/county names, FIPS codes, abbreviations)
- **Support for multiple crop types** (rice varieties, chickpea sizes, etc.)
- **Historical program year handling** (automatic SRP/ERP selection)
- **Custom price data support** for specialized analyses
- **Quiet mode** for batch processing

### Key Parameters

- `crop`: The crop name (e.g., "corn", "soybeans", "rice")
- `crop_type`: Specific variety for crops with multiple types
- `program_year`: The program year for calculation
- `base_acres`: Number of enrolled acres
- `mya_price`, `erp`, `srp`, `nmlr`: Price parameters (optional, uses FSA data if not provided)
- `plc_yield`: Yield parameter (optional, calculated from location if not provided)
- `state`, `county`, `fips`: Location parameters for yield calculation
- `cov_lvl`: Coverage level (default 0.85)
- `quiet`: Suppress messages for batch processing

For more detailed information about specific parameters and options, see the function documentation with `?calc_plc_payment`